---
title: "SPP608 Statistical Methods For Public Policy Lab 2"
author: "Viviana Wu"
date: "2/11/2021"
output:
  pdf_document:
    toc: yes
    toc_depth: '4'
  html_document:
    highlight: tango
    theme: lumen
    toc_float: yes
---


```{r, include=FALSE}
options(tinytex.verbose = TRUE)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Lecture

Slides for Week 2 is available [here](https://docs.google.com/presentation/d/1vIfUgwOKx0wVSsujtuyLB6OFf6OqYxnw1gfLlpoH-NA/edit?usp=sharing)

In this tutorial, you will learn how to:

* use import, explore and visualize a dataset
* understand a data structure of variables in R 
* make decisions to clean the data 

## Step 1: Import a Dataset

'data' is the variable where the data will be stored.  If the parameter "header=" is "TRUE", then the first row will be treated as the row names. These data were reported in Almond, D., Chay, K. Y., & Lee, D. S. (2005). The costs of low birth weight. *The Quarterly Journal of Economics*, 120(3), 1031-1083. which is made available via Stock, J. H., & Watson, M. W. (2020). *Introduction to Econometrics.* 4th ed. Pearson. NY: New York. 
```{r}
data <- read.csv("birthweight_smoking.csv",header=TRUE, sep=",")

# Now on your right panel in your global environment, you should see 'data' stored. It has 3,000 observations (i.e., 3,000 rows) and 12 variables (i.e., 12 columns). 

```

Desktop version (We'll walk through next week): If your data is not stored in the project file, you will need to insert the full path for your data, remember to use the forward slash '/" 


```{r}
# data <- read.csv("C:/Users/bijou/OneDrive - University of Massachusetts/SPP Stats 608 2021/Labs/Data sets/birthweight_smoking.csv",header=TRUE, sep=",") 

```

What if our data was saved as an Excel sheet (.xlsx)? We first install a new package called "readxl": **install.packages("readxl")** and run the following codes:
```{r}
# install.packages("readxl")
# library(readxl)
# data <- read_excel("birthweight_smoking.xlsx") 
```
Alternatively, you can open it through the files pane (in the lower right panel) by clicking the .xlsx file

***

## Step 2. Exploratory Data Analysis  

### dim()   
Examine the dimensions of your dataset, it returns two numbers: (1) # of Rows (2) # of Columns.	
```{r}
dim(data)
```

### options(scipen=999)
Remove the scientific notation in the format of our data.
```{r}
options(scipen=999)
```

### colnames()
What are the variables included in the dataset?
```{r}
colnames(data)
```

### head()
`head()` shows the first six rows of a dataset. 
```{r}
head(data)
```


### str()
Examine the data structure of the variables in the data frame (factor,numeric,integer,etc.).
```{r}
str(data)
# want to find out the type of data structure for a particular variable? Try it yourself.

```

### summary()
Examine the summary statistics of variables in the dataset, what can you learn from it?
Important: Check against the "Documentation for Birthweight_Smoking.pdf" to understand the meaning of the variable and the labeling system. 

```{r}
summary(data)

```

We can take a closer look into specifc variables in the `data`. We use `$` to indicate that we are looking for a variable of interest within `data`. For instace, we want to find out the data structure and summary information ofthe  `smoker` variable within the `data` dataset. 
```{r}
str(data$smoker)
summary(data$smoker)

# What is the data structure of `data$smoker`?

```

### Q1: Can you repeat the above example for **a continuous variable** and **a dummy variable** in the dataset? 
 > Remember, R is case-sensitive, needs to be an exact match. You can copy the variable names from summary result. 

Tips: A dummy variable has dichotomous values, e.g., 0 or 1, 1 or 2, etc. A continuous variable has continuous numerical values.

A Continuous variable: ________
```{r}
# use str() and summary()

```

Describe the mean, median, minimum and maximum of the variable here: __________

  
A Dummy variable: _________
```{r}
# use str() and summary()


```
Describe the minimum and maximum of the dummary variable here: __________


### rename_at() from library(dplyr) 
If you use a desktop version, you might have this naming problem. 
To rename a variable `ï..nprevist`, we can use the function `rename_at()` from `library(dplyr)`.  
```{r , eval=FALSE}
# install.packages("dplyr")
library(dplyr)
data <- data %>% rename_at("ï..nprevist",~"nprevist")  # rename_at("old name), ~"new name")
summary(data)
# A initutive but two step approach: 
    # data$nprevist <- data$ï..nprevist  # create a new variable from data$ï..nprevist 
    # data$ï..nprevist <- NULL # remove this variable
    # str(data)
```

***   

## Step 3: Data Cleaning

Before proceeding to analyze our data, we have to do some data cleaning. 

### as.factor() & assign labels

For a dummy variable, we need to change the structure from an integer to a factor (categorical/dummy) variable in R. 

For categorical variables, we can only count the occurrence in each categories, unlike for continuous variables we can measure the values and measure the distribution (mean, median, variance, etc.). 
```{r}
# change the data structure to a factor variable in R
data$smoker <- as.factor(data$smoker)   

# Now, check if the data structure for `smoker` has changed 
str(data$smoker)  

# See how the summary statistics has changed. 
summary(data$smoker)  

# We can **assign value labels** to the variable `smoker` (0 = no, 1=yes).
data$smoker <- factor(data$smoker,levels = c(0,1),labels = c("no", "yes"))
summary(data$smoker)

```

### Q2: Change the data type of another dummy variable, and assign the value labels. 

Dummy variable: _______
```{r}



```


Let's **create a new variable** for birthweight in oz (instead of grams) named `birthweight_oz` by multiplying `birthweight` by 0.035274. (Grams to Ounces formula: oz = g * 0.035274)
```{r}
data$birthweight_oz <- data$birthweight * 0.035274
summary(data$birthweight_oz)
```

### NULL 

Sometimes, you want to **delete** a variable. For instance, you can remove the variable `no_previst` from the `data` as follows: 
```{r}
data$no_previst<-NULL
```

### subset()

**Subset the data** by a specific family id number by the family id number 16922, by the value of home94 equal to 1, and lastly, adding an additional condition that NFA94 has a value greater than 10000. 
```{r}
summary(data)
data_smoker <- subset(data, data$smoker == 1)
data_adult <- subset(data, data$age >= 18) 
data_smoker_adult <- subset(data, data$smoker == 1 & data$age >= 18) 
```
Try it yourself!

###  Q3: Subset the data and create a new dataframe, choose any critera!

```{r}
# data_new <- subset(data, data$____ (an operator) ___ )


```

***  

## Step 4: Data Visualization

### table() and prop.table()

Create a **frequency table** of a factor variable; The frequencies are ordered and labeled by the levels attribute of the factor.
```{r}
table(data$smoker)   # How many 0s and 1s in `smoker`?

# let's label the value!
data$smoker <- factor(data$smoker,levels = c(0,1),labels = c("no", "yes")) 

table(data$smoker) # you can see the label now.
```

We can express table as proportion tables or fraction Of marginal table of two variables. 
```{r}
# for one variable
round(prop.table(table(data$unmarried)),2) 
round(prop.table(table(data$educ)),2)

# for two variables
prop.table(table(data$unmarried, data$smoker), margin=2)   # row = unmarried, column = smoker
```

### hist()

Create a histogram for the continuous variable (also our dependent variable) `birthweight` and examine the distribution, how would you describe the distribution?
```{r}
hist(data$birthweight)
```

### ggplot(): Histogram

Create a histogram and a scatterplot, what can you observe from the graph?
```{r warning=FALSE, message=FALSE, eval=TRUE}
# install.packages("ggtheme")
# install.packages("ggplot2")
library(ggplot2)
library(ggthemes) 

ggplot(data, aes(birthweight)) +
  geom_histogram() 

# For histograms with tick marks between each bin, use `geom_bar` + `scale_x_binned`.
ggplot(data, aes(birthweight)) +
  geom_bar() +
  scale_x_binned()
```


### Q4: Use `ggplot()` to create a histogram for another continuous variable other than `birthweight`

```{r}



```


### ggplot(): Scatterplot 

```{r warning=FALSE, message=FALSE , eval=TRUE}

plot <- ggplot(data, aes(x=nprevist, y=birthweight)) + geom_point() +
  stat_smooth(method="lm", se=FALSE)

plot 

# can add themes  

plot + labs(title = "Scatterplot of prenatal visits and birth weight of infant", 
            subtitle = "Lab 2") + labs(caption = "(based on data in Pennsylvania in 1989)") +
       geom_point(color='lightblue4') +
       geom_smooth(method="lm", colour="black", size=0.5) + 
       labs(x = "Number of Prenatal Visits", y = "Infants' birthweight (in grams)") +
       theme_economist(base_size = 10, base_family = "sans") 
  

  # Other themes e.g., theme_grey, theme_bw, theme_linedraw, theme_light, theme_dark, theme_minimal or theme_classic
  # customize your theme
  # https://www.r-bloggers.com/custom-themes-in-ggplot2/ 
  # https://ggplot2.tidyverse.org/reference/geom_histogram.html 


```

### Q5: Plot a new scatterplot using `ggplot()` with a new independent variable (x)
Feel free to add themes, labels, colors that make your graph look AWESOME!
Tips: Continuous variables will have a "nicer" visualization 

```{r}
# plot_new <- ggplot(data, aes(x= _____ , y=birthweight)) + geom_point() +
#  stat_smooth(method="lm", se=FALSE)

# plot_new 


```


### R Operators

R has several operators to perform tasks including arithmetic, logical and bitwise operations.

| Operator | Meaning |
|-------- | ------- |
| <  | less than |
| <=	| less than or equal to |
| >	| greater than |
| >=	| greater than or equal to|
| ==	|exactly equal to |
|  !=	|not equal to |
| !x	|Not x |
| x | y |x OR y |
| x & y	|x AND y |
| isTRUE(x)	|test if X is TRUE |


***

## SELF-GRADING (upon receiving the answer key)

 * Step 1: Check if you got the correct answers for all questions. 
 
  **All correct:**  
  Type the grading of your homework and do a short recap of what you learn:
  e.g., All Correct, statastic!!! I have learnt... 
  
  **Some mistakes:**
  Explain the mistakes made and re-do the coding in the code chunk below. 

```{r}
# provide question number and re-run codes here. 




```

 * Step 2: `Knit` your R Markdown document--move your cursor to the upside-down triangle next to `Knit`, and choose for **PDF**.  

 * Step 3: Submit this updated copy of R Markdown PDF to Gradescope **Lab 2 Assignment (Self-graded)** <https://www.gradescope.com/>

*** 

